---
title: "UD651 - Problem Set 3"
author: "Ivan Heneghan"
date: "8 January 2016"
output: html_document
---

This is the project report for UD651 - Problem Set 3.

Install all relevant packages

```{r}
install.packages("ggplot2", repos = "http://cran.us.r-project.org")
install.packages("gridExtra", repos = "http://cran.us.r-project.org")
library(ggplot2)
library(gridExtra)
library(xlsx)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr, warn.conflicts = FALSE)
```

Load diamonds data set, and perform initial analysis

```{r}
str(diamonds)
head(diamonds)
colnames(diamonds)
summary(diamonds)
View(diamonds)
```

Produce histogram of price of all the diamonds

```{r, echo=FALSE}
qplot(x = price, data = diamonds, binwidth = 10)
```

Summary of price information

```{r}
summary(diamonds$price)
```

Counts

```{r}
sum(diamonds$price < 500)
sum(diamonds$price < 250)
sum(diamonds$price >= 15000)
```

Exploring the largest peak in the histogram

Entire histogram

```{r}
qplot(x = price, data = diamonds, binwidth = 250, 
      xlab = "Price", ylab = "Count of diamonds") +
  scale_x_continuous(limits = c(250, 20000), breaks = seq(250, 20000, 2500))
```

Zooming in on highest peak
```{r}
qplot(x = price, data = diamonds, binwidth = 50, 
      xlab = "Price", ylab = "Count of diamonds") +
  scale_x_continuous(limits = c(250, 2000), breaks = seq(250, 2000, 250))
```

Break out prices by cut

Histograms

```{r}
qplot(x = price, data = diamonds, binwidth = 250, 
      xlab = "Price", ylab = "Count of diamonds") +
  scale_x_continuous(limits = c(250, 20000), breaks = seq(250, 20000, 2500)) + 
  facet_wrap(~cut)
```

Numerical analysis

```{r}
by(diamonds$price, diamonds$cut, summary)
```

Using Max due to rounding

```{r}
by(diamonds$price, diamonds$cut, max)
```

Scales & multiple histograms

```{r}
qplot(x = price, data = diamonds, binwidth = 10) + 
  facet_wrap(~cut, scales = "free_y")
```

Price per Carat by Cut

```{r}
qplot(x = price/carat, data = diamonds, binwidth = 50) +
  facet_wrap(~cut, scales = "free_y") +
  scale_x_log10()
```

Box plot - Price of diamonds, numerical summaries, and one other variable

```{r}
qplot(x = color, y = price, 
      data = diamonds, 
      geom = "boxplot",
      color = color) +
  coord_cartesian(ylim = c(250, 20000))
```

IQR analysis

```{r}
by(diamonds$price, diamonds$color, summary)
IQR(subset(diamonds, color == "D" )$price)
IQR(subset(diamonds, color == "J")$price)
```

Box plot - Price per carat of diamonds across colors

```{r}
qplot(x = color, y = price/carat , 
      data = diamonds, 
      geom = "boxplot",
      xlab = "Color",
      ylab = "Price per Carat",
      color = color) +
  coord_cartesian(ylim = c(250, 20000))
```

Frequency polygon

```{r}
qplot(x = carat, data = diamonds, 
      binwidth = 0.05,
      xlab = "Carat",
      ylab = "Count",
      geom = "freqpoly") +
  scale_x_continuous(limits = c(0, 5), breaks = seq(0, 10, 0.1))
```

Gapminder analysis - working hours per week

1. Import data

```{r}
whpw <- read.xlsx("D:\\Foundations of Data Science\\Projects\\EDA - Facebook\\fods-ud651-ps3\\indicator_hours per week.xlsx", 1, stringsAsFactors = FALSE)
```

2. Rename first column to something more meaningful

```{r}
names(whpw)[1] <- c("country")
```

3. Tidy data - gather into two column key-value pair

```{r}
whpw <- whpw %>% gather(year, hours, -country)
```

4. Remove X's from start of years

```{r}
whpw$year <- gsub("X", "", whpw$year)
```

5. Create data table from the imported data

```{r}
tbl_whpw <- tbl_df(whpw)
```

6. Initial viewing of data

```{r}
str(tbl_whpw)
head(tbl_whpw)
colnames(tbl_whpw)
summary(tbl_whpw)
View(tbl_whpw)
```

7. Analysis - boxplot showing progression over time

```{r}
ggplot(tbl_whpw[!is.na(whpw$hours),], aes(x = year, y = hours, fill = year)) + 
  geom_boxplot() +
  xlab("Year") +
  ylab("Working hours per week") +
  ggtitle("Working hours per week over time")
```

8. Analysis - line graph

```{r}
ggplot(aes(x = year, y = hours, group = country, color = country), data = tbl_whpw, binwidth = 5) +
  geom_line() +
  xlab("Year") +
  ylab("Working hours per week") +
  ggtitle("Working hours per week over time")
```

9. Analysis - subsetting to look at countries with > 40 hours per week over entire time period

```{r}
ggplot(aes(x = year, y = hours, group = country, color = country), data = subset(tbl_whpw, hours > 40) , binwidth = 5) +
  geom_line() +
  xlab("Year") +
  ylab("Working hours per week") +
  ggtitle("Working hours per week over time")
```

10. Analysis - subsetting to look at countries with < 40 hours per week over entire time period

```{r}
ggplot(aes(x = year, y = hours, group = country, color = country), data = subset(tbl_whpw, hours < 40) , binwidth = 5) +
  geom_line() +
  xlab("Year") +
  ylab("Working hours per week") +
  ggtitle("Working hours per week over time")
```

11. Melting data to allow delta change calculation & histogram

```{r}
tbl_whpw <- spread(tbl_whpw, year, hours)
```

12. Getting delta change for each country between 1990 and 2007

```{r}
tbl_whpw$delta <- tbl_whpw$`2007` - tbl_whpw$`1990`
```

13. Analysis - Reviewing sumary of delta change

```{r}
summary(tbl_whpw$delta)
```

14. Analysis - looking at histograms for 1990 and 2007
```{r}
g1 <- ggplot(aes(tbl_whpw$`1990`), data = tbl_whpw, binwidth = 1) + 
  geom_histogram() +
  xlab("Working hours per week") +
  ylab("Count of countries") +
  ggtitle("Working hours in 1990")

g2 <- ggplot(aes(tbl_whpw$`2007`), data = tbl_whpw, binwidth = 1) + 
  geom_histogram() +
  xlab("Working hours per week") +
  ylab("Count of countries") +
  ggtitle("Working hours in 2007")

grid.arrange(g1, g2, ncol = 1)
```
